Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-22 draft.tex
Wrote on Application Section

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{Interactive Visual Analysis, Epidemiology}

@@

Designing a visualization which communicates all aspects of the data equally is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, it is often a good solution to combine the strength of different visualization techniques in a unified system \cite{Buja91, Konyha2009}.
%
The Principal Component Analysis (PCA) and similar techniques are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to determine.
\\\\
The work of Turkay and colleagues is closest to ours albeit our focus on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
%
Investigating Data on an norwegian aging study their methods aim to amplify a hypothesis generation process.
%
Statistical measures of metric variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots}.
%
These transform dimensions into data points and make them comparable with respect to the derived descriptive measures.
%
This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
The technique was applied to variables generated by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive measures of the epidemiological factors.
%
Hypothesis based on observations of changes in these plots may impose over-fitting to the data because the measure highlights only subsets of statistical changes.
%
Our approach sticks more to the information extracted from the segmented image data and derive variable interaction with non-image epidemiological factors.
% Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!

@@

\section{Interaction- and Visualization Techniques}
% TODO: Bernhard: Section anders beginnen!
We employ the fourth and highest level which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
\begin{figure}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/visualization}
 \caption{\emph{Image is not final} Screenshot from the front-end which is divided as follows: (a) Sidebar which contains all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive Pivot Table which shows exact numbers for each displayed variable combination. The data displayed is used to analyze the lumbar spine.
 }
 \label{fig:visualization}
\end{figure}
\emph{IVA} levels define different levels interaction. 
%
The employed techniques needs to respect the epidemiological requirements.
%
Hypothesis generation bears the chance of over-fitting the data to expectations.
%
To avoid this, a timeline needs to be introduced which keeps track how many variable variations the user evaluated before coming to an conclusion.
%
Since epidemiologists are used to process groups based on table representations we decided to introduce an interactive solution in form of a Pivot Table.

@@

\subsection{Automated Feature Suggestion}
As discussed previously, highlighting potential interesting values in the data set is one major benefit of the \emph{IVA} powered approach.
%
Turkay and colleagues used the approach to calculate various key figures based in the distribution functions of each feature derived from the image data \cite{Turkay2013}.
%
Since the majority of our data are categorical features, we have to employ different solutions.
%
A solution for this problem are odds ratios, which are a standard statistical tool for stating relations between features.
%
Odds ratios can only be calculated for $2\times2$ contingency tables which usually represent the presence of a condition in a population divided by a characteristic (e.g. male/female subjects with presence or absence of back pain).
%
To calculate odds ratios for variables with more than two manifestations, we calculate local odds ratios for each possible combination, yielding a matrix for each feature combination \cite{Rudas1998}.
%
When the subjects are divided into groups and the calculation is carried out again for all feature combination, the difference in sum of the odds weighted with the number of manifestations per feature can be used to indicate if the feature combination yields a difference.
%
These difference are then highlighted in a separate tab of the side bar "Interactions".
%
Interesting interactions then can be assessed creating a linked view using the standard drag and drop workflow.
\\\\
In the following sections we will discuss details on the implementation which relies on modern Web-Technologies.
\\\\
ToDo\\
- This can be improved--summing up the values possibly not the cleverest solution--calculation of variance etc. possible\\
- Matrix Visualization?

@@

\section{Application}
We apply the presented set of techniques to a data set which is compiled to analyze lower back pain which is one of the most common reasons for an adult to see a physicians \cite{Backpain}.
%
Data confidentiality and ethical reasons prohibit us from accessing the complete data space of the SHIP feature space.
%
The epidemiologists compiled feature list which is a tradeoff between complexity and ability to pass the ethics committee.
%
%The goal is to find features which \emph{interact} with lower back pain based on image and non-image data.
%
\subsection{The Spine Dataset}
We divide the data set in image and non-image data.
%
There are 136 features describing diagnosed diseases, lifestyle factors, women specific factors, pain indicators, laboratory values somatometric variables and are ordered accordingly.
%
The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
%
The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.

\subsection{Processing the Image data}
The lumbar spine was detected in the image data using a hierarchical finite element method according to \cite{Rak2013}.
%
This semi-automatic method requires the user to initialize the Tetrahedron-based finite element models (FEM) with a click on the L2-vertebra.
%
It uses a weighted sum of T1- and T2-weighted MRI images to detect the lumbar spine shape.
%
The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1 -L5 vertebrae.
%
For clustering purposes we extracted the centerline of the spine canal of the lumbar spine canal which captures information about lordosis and scoleoses which is the medical term for spine curvature \cite{Klemm2013VMV}.
%
%This information is a major feature for diagnosing lumbar back pain.
\subsection{Processing the Non-Image data}
- SPSS Export? JSON-Export? Automatisches Anfuegen des Dictionaries?

%\subsection{The Spine Prototype}

- Clustering using Steffens Algorithm
% - Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
% - Problem of sparse differences - Visualization has to be more abstract to emphasize the differences.\\
% - Input of Epidemiologists goes here!\\
% \\\\
% \textbf{From VMV'13 Paper}\\
% 
% All whole-body MRI scans were acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way. Subjects were placed in the supine position. Five phased-array surface coils were placed to the head, neck, abdomen, pelvis, and lower extremities for whole-body imaging. The spine coil is embedded in the patient table. The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels). First, both sequences were placed over the cervical and upper thoracic spine. Then, they were placed over the lower thoracic and lumbar spine. The MRI software automatically composed a whole spine sequence from the two T1-weighted and T2-weighted sequences \cite{Hegenscheid2013}. We were provided with 490 data sets.

@@

\section{Summary and Conclusion}
- Future Work: Matrix-View of differences - more intuitive\\
- UI more flexible by hiding UI-panes

>>>>

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{Interactive Visual Analysis, Epidemiology, Spine}

@@

Designing a visualization which communicates all aspects of the data equally is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, it is often a good solution to combine the strength of different visualization techniques in a unified system \cite{Buja91, Konyha2009}.
%
The Principal Component Analysis (PCA) and similar techniques are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to determine.
\\\\
The work of Turkay and colleagues is closest to ours albeit our focus on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
%
Investigating Data on an norwegian aging study their methods aim to amplify a hypothesis generation process.
%
Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots}.
%
These transform dimensions into data points and make them comparable with respect to the derived descriptive measures.
%
This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
The technique was applied to variables generated by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive measures of the epidemiological factors.
%
Hypothesis based on observations of changes in these plots may impose over-fitting to the data because the measure highlights only subsets of statistical changes.
%
Our approach sticks more to the information extracted from the segmented image data and derive variable interaction with non-image epidemiological factors.
% Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!

@@

\section{Interaction- and Visualization Techniques} \label{Interaction- and Visualization Techniques}
% TODO: Bernhard: Section anders beginnen!
We employ the fourth and highest level which includes next to brushing and linking, advanced brushing, extraction of attributes the development of visualizations custom tailored to the data sets.
\begin{figure}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/visualization}
 \caption{\emph{Image is not final} Screenshot from the front-end which is divided as follows: (a) Sidebar which contains all features as well as the groups defined in the analysis process; (b) Canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) Interactive Pivot Table which shows exact numbers for each displayed variable combination. The data displayed is used to analyze the lumbar spine.
 }
 \label{fig:visualization}
\end{figure}
\emph{IVA} levels define different levels interaction. 
%
The employed techniques needs to respect the epidemiological requirements.
%
Hypothesis generation bears the chance of over-fitting the data to expectations.
%
To avoid this, a timeline needs to be introduced which keeps track how many variable variations the user evaluated before coming to an conclusion.
%
Since epidemiologists are used to process groups based on table representations we decided to introduce an interactive solution in form of a Pivot Table.

@@

\subsection{Automated Feature Suggestion}
% https://de.wikipedia.org/wiki/Kontingenzkoeffizient#Cram.C3.A9rs_V
TODO REWRITE USING CRAMERS V

As discussed previously, highlighting potential interesting values in the data set is one major benefit of the \emph{IVA} powered approach.
%
Turkay and colleagues used the approach to calculate various key figures based in the distribution functions of each feature derived from the image data \cite{Turkay2013}.
%
Since the majority of our data are categorical features, we have to employ different solutions.
%
A solution for this problem are odds ratios, which are a standard statistical tool for stating relations between features.
%
Odds ratios can only be calculated for $2\times2$ contingency tables which usually represent the presence of a condition in a population divided by a characteristic (e.g. male/female subjects with presence or absence of back pain).
%
To calculate odds ratios for variables with more than two manifestations, we calculate local odds ratios for each possible combination, yielding a matrix for each feature combination \cite{Rudas1998}.
%
When the subjects are divided into groups and the calculation is carried out again for all feature combination, the difference in sum of the odds weighted with the number of manifestations per feature can be used to indicate if the feature combination yields a difference.
%
These difference are then highlighted in a separate tab of the side bar "Interactions".
%
Interesting interactions then can be assessed creating a linked view using the standard drag and drop workflow.
\\\\
In the following sections we will discuss details on the implementation which relies on modern Web-Technologies.
\\\\
ToDo\\
- This can be improved--summing up the values possibly not the cleverest solution--calculation of variance etc. possible\\
- Matrix Visualization?

@@

\section{Application}
We apply the presented set of techniques to a data set which is compiled to analyze lower back pain which is one of the most common reasons for an adult to see a physicians \cite{Backpain}.
%
Epidemiological analysis of lumbar back pain such the work of Harreby and colleagues \cite{Harreby1996} is largely focused on non-image information.
%
If at all, only a few shape related features are included in comparable studies \cite{Lang2011}.
%
To our knowledge, this is the first approach on analyzing shape related information of the whole lumbar spine with other epidemiological features.

%
Determining risk factors can lead to \cite{Fletcher2012}:
\begin{itemize}
	\item a better understanding of effects of preventive measures such as occupational health and safety regulations
	\item prognostic features for diagnosis and treatment of lumbar back pain
	\item determination of particularly effected risk groups
\end{itemize}
%
Characterizing the healthy aging process of the spine is a large stretch goal that allows to determine age-normalized probabilities for diseases using individual risk factors.

Data confidentiality and ethical reasons prohibit us from accessing the complete data space of the SHIP feature space.
%
The epidemiologists compiled feature list which is a tradeoff between complexity and ability to pass the ethics committee.
%
%The goal is to find features which \emph{interact} with lower back pain based on image and non-image data.
%
\subsection{The Lumbar Spine Dataset}
We divide the data set in image and non-image data.
%
There are 136 features describing diagnosed diseases, lifestyle factors, women specific factors, pain indicators, laboratory values somatometric variables and are ordered accordingly.
%
The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
%
The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.

\subsection{Data Preprocessing} \label{Data Preprocessing}

\paragraph{Image-Data.} \label{Image-Data}
The lumbar spine was detected in the image data using a hierarchical finite element method according to \cite{Rak2013}.
%
This semi-automatic method requires the user to initialize the Tetrahedron-based finite element models (FEM) with a click on the L2-vertebra.
%
It uses a weighted sum of T1- and T2-weighted MRI images to detect the lumbar spine shape.
%
The registered models capture resilient information about shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae.
%
For clustering purposes we extracted the centerline of the spine canal of the lumbar spine canal which captures information about lordosis and scoliosis which are medical terms for spine curvature \cite{Klemm2013VMV}.

TODO HOW MANY SUBJECTS?

\paragraph{Non-Image Data.}

In order to allow for fast and easy access of the data outside of tools like \texttt{SPSS} or \texttt{STATA} the data was exported to a JSON file format which can easily be loaded using Javascript.
%
Each variable is stored as object which contains: 
\begin{itemize}
	\item the data as array of values--categorical values and error codes are stored using IDs
	\item the data type (continuous, nominal, ordinal, dichotomous)
	\item a detailed description of the variable
	\item the data dictionary which translates value- or error IDs to the actual values
\end{itemize}
%
% Binning: https://www.researchgate.net/post/What_metric_best_captures_the_information_loss_when_binning_a_large_data-set_into_a_histogram_representation
Continuous variables are discretized to allow for CramérsV contingency coefficient assessment.
%
In epidemiology, continuous data is usually categorized into ordinal groups of equal size.
%
Since the number of categories often strongly depends on the hypothesis, the discretization steps can be adapted dynamically.
%
To allow for hypothesis generation we set the number of groups to five if not specified otherwise.
TODO Formula?

\subsection{Shape Visualization and Clustering}
The detection algorithm described in Section~\ref{Data Preprocessing} yields a model that determines voxel belonging to the lumbar spine.
%
The tetrahedron-based model has consists of corresponding grid points for each structure instance.
%
This allows for calculation of shape-distance and similarity.
%
We use this information to calculate mean-shapes for groups for plotting them as described in Section~\ref{Interaction- and Visualization Techniques}.
%
For dichotomous variables the color codes distances between mean shapes of the two groups, for variables with more than two manifestation the color encodes the distance to the mean shape of all variables.
\\\\
Shape based clustering is carried out via Agglomerative Hierarchical Clustering of the spine canal centerlines that are described in Section~\ref{Data Preprocessing} \cite{Klemm2013VMV}.
%
Since it is not possible to determine the number of clusters in a given group, it is determined automatically by computing a \emph{knee/elbow} point which is described as good tradeoff between number of cluster and a cluster evaluation metric \cite{Salvador2004}.
%
For details, see \cite{Klemm2013VMV}.
%
The method has proven to produce comprehensible results on a preliminary data set and was able to reproduce textbook knowledge before \cite{Klemm2013VMV}.

\subsection{Exploratory Analysis of the Lumbar Spine Dataset}
%
We analyzed the given dataset expert guided to assess its suitability for both hypothesis validation and generation.
%
\subsubsection{Hypothesis free Analysis}
\begin{figure}[htb]
 \centering
 \label{fig:hypopthesisfree}
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/hypothesisfree}
 \caption{\emph{Image is not final}. (a) Clustering result of all subjects. The bar chart height indicates the number of subjects in the cluster. The difference to the mean-shape is color-coded where blue represents no difference and red a large difference.}
\end{figure}
%
Analyzing the data set without a prior hypothesis requires a starting point which gives a overview over the data first \cite{Shneiderman1996}.
%
Shape based hypothesis free exploration starts with a shape-grouping step achieved by image-based clustering.
%
The results for this step can be seen in Figure~\ref{fig:hypopthesisfree}.
%
Cluster 9 represents the subjects with average shape.
%
Other shapes differ with respect to size, such as Cluster 2, 8, 10 and 3 where the last one also is more straight, which is expected for larger subjects.
%
Cluster 4, 7 and 5 contain outliers, characterized by their unusual shape and small number.
%
Cluster 8 - Viele nicht erwerbstätige - weiter hier

TODO: While this approach does not assume any hypothesis, the data is met with a selection bias when compiling the list of related features by the domain experts.

TODO Implement Color coding only in a certain direction!

\subsubsection{Hypothesis based Analysis}
%

%

\subsubsection{Concluding Domain-Expert Feedback}
Our clinical partners were interested in the MRI data for the subjects in the outlier clusters because they are highly likely to exhibit pathologies.
%
We plan to include a \texttt{DICOM}-viewer to meet this wish.

% - Describe steps from gathering Information from the raw image files (segmentation, abstraction, visualization)\\
% - Problem of sparse differences - Visualization has to be more abstract to emphasize the differences.\\
% - Input of Epidemiologists goes here!\\
% \\\\
% \textbf{From VMV'13 Paper}\\
% 
% All whole-body MRI scans were acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way. Subjects were placed in the supine position. Five phased-array surface coils were placed to the head, neck, abdomen, pelvis, and lower extremities for whole-body imaging. The spine coil is embedded in the patient table. The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels). First, both sequences were placed over the cervical and upper thoracic spine. Then, they were placed over the lower thoracic and lumbar spine. The MRI software automatically composed a whole spine sequence from the two T1-weighted and T2-weighted sequences \cite{Hegenscheid2013}. We were provided with 490 data sets.

@@

\section{Summary and Conclusion}
- Future Work: Matrix-View of differences - more intuitive\\
- UI more flexible by hiding UI-panes
- Dynamic Discretization to better fit the data distributions
- More Shape based filtering - reorder by size, curvature, etc. ...
- Color code different Aspects - Only difference in x or y direction

Discretization of continous variables into bins of equal size may bias the underlying data by not considering the variable distribution function \cite{Fletcher2012}.
% Fletcher 2012 - S. 257
To reduce the number of false positive findings, the data space can be randomly be cut in half the hypothesis then can be cross-validated for statistical soundness.
%
This requires a large number of subjects, especially if the investigated features are rare and are only presented by a few subjects.
%
As suggested by Fletcher and Fletcher, we plan to validate the hypothesis using the \texttt{SHIP-TREND} cohort \cite{Fletcher2012}.
